# Exercise 2: Identifying letters of alphabet

Wolford and Hollingsworth (1974) were interested in the confusions made when a person attempts to identify letters of the alphabet viewed for some milliseconds only. A confusion matrix was constructed that shows the frequency with which each stimulus letter was mistakenly called something else. A section of this matrix is shown in the table below.

```{r, warning=FALSE, echo=FALSE}
library(knitr)
library(kableExtra)
df <- data.frame(Letter=c('B','C','D','F','G'),
                 B=c('--','3','7','3','7'),
                 C=c('','--','5','5','12'),
                 D=c('','','--','7','2'),
                 F=c('','','','--','2'),
                 G=c('','','','','--'))
kable(df, escape=F)
```

The dataset is available from 'Datasets for week 2', under the name 'letter.csv'. The task is to visualise the letters as points in one/two/three dimensions and discover if there is anything interesting. 

**Task 2**

(a) Is it appropriate to use principal component analysis to visualise the data? `r mcq(c("Yes", answer="No"))`

(b) Is this dataset a dissimilarity matrix? `r mcq(c("Yes", answer="No"))`  

(c) Apply an appropriate MDS method to this data. 

(d) How many dimensions would you keep?

(e) Comment on any pattern from the configuration plot. 

`r hide("Hint")`
If you would like to visualise the data in 3D, the following commands may be helpful. 
```{r, eval=FALSE}
library(rgl)
plot3d(...)
text3d(...)
```
`r unhide()`

(f) Apart from the number of dimensions, have you introduced any other parameter(s) in your analysis? If so, study their influence by trying different values.

(g) Finally, as mentioned in the lecture note, MDS are sensitive to initial configurations. Try multiple initial configurations and compare the results. 
*You may need to look into the help page of `mds` to find out how to change the initial configuration.*
